Dissertation

### 2024-04-12 Sections/full.qmd
update/re-render all versions. Took authors note out of hikmah manuscript

---
title: Full Dissertation
subtitle: "The Role of Variability in Learning Transfer: A Similarity-Based Computational Approach"
date: 04/04/24
author:
- name: Thomas Gorman
  affiliation: Indiana University
page-layout: full
code-fold: true
code-tools: true
toc: true
toc-location: body
toc-title: Contents
cache: true
execute: 
  warning: false
  eval: true
format:
  html: default
  hugo-md:
    echo: false
    #include: false
    html-math-method: mathjax
    output-file: diss_full_hugo.md
  gfm:
    echo: false
    output: false
    output-file: diss_full_gfm.md
  hikmah-manuscript-pdf:
    echo: false
    output-file: "tg_full_disser.pdf"
    mainfont: "Linux Libertine O"
    mainfontoptions:
      - "Numbers=Proportional"
      - "Numbers=OldStyle"
    mathfont: "Libertinus Math"
editor: 
  markdown: 
    wrap: 72
prefer-html: true
---

@@

Early models of learning often assumed that discrete experiences with some task or category were not stored individually in memory, but instead promoted the formation of a summary representation, often referred to as a prototype or schema, and that exposure to novel examples would then prompt the retrieval of whichever preexisting prototype was most similar. In addition to being a landmark study on the influence of training variability, @posnerGenesisAbstractIdeas1968 (described above) also put forward an influential argument concerning the nature of the mental representations acquired during learning - namely that learners tend to abstract a prototype, or aggregate representation of the dot pattern categories, rather than encoding each individual stimuli. Recall that participants are trained on only on distortions of the category prototypes (e.g.  low, medium or high distortions), never encountering the exact prototypes during the training stage. Then, in the testing phase, participants are tested with the prototype patterns, their old training items, and novel low, medium and high distortions. The authors found that participants had the highest testing accuracy for the previously unseen prototype patterns, followed by the old training items, and then the novel low, medium and high distortions. The authors interpreted this pattern as evidence that participants had acquired prototype representation of the category, as opposed to storing each individual training instance, and that generalization was based on the similarity of the testing items to the learned prototype representations. @posnerGenesisAbstractIdeas1968 has been extremely influential,  and continues to be cited in contemporary research as clear evidence that prototype abstraction underlies the benefits of varied training. It's also referenced as a key influence in the development of "Schema Theory of Motor Learning" @schmidtSchemaTheoryDiscrete1975, which in turn influenced decades of research on the potential benefits of varied training in motor skill learning. However a number of the core assumptions utilized by @posnerGenesisAbstractIdeas1968 were later called into question both empirically and with competing theoretical accounts [@hintzmanMINERVASimulationModel1984; @hintzmanSchemaAbstractionMultipletrace1986; @palmeriCentralTendenciesExtreme2001; @zakiHighdistortionEnhancementEffect2007; @mcclellandDistributedMemoryRepresentation1985; @nosofskyInvestigationsExemplarBasedConnectionist1992]. @palmeriCentralTendenciesExtreme2001 demonstrated the both the dangers of assuming that psychological representations mimic the metric stimulus space, as well the viability of models with simpler representational assumptions. These authors conducted a near replication of the @posnerGenesisAbstractIdeas1968 study, but also had participants provide similarity judgements of the dot pattern stimuli after completing the training phase. A multidimensional scaling analysis of the similarity judgements revelead that the psychological representations of the prototype stimuli were not located in the middle of the training stimuli, but were instead extreme points in the psychological space. The authors also demonstrated the generalization patterns of @posnerGenesisAbstractIdeas1968 could be accounted for by an exemplar-based model, without any need to assume the abstraction of a prototype.

@@

_______________________________________________________________________________

@@

The study of human function learning investigates how people learn relationships between continuous input and output values.  Function learning is studied both in tasks where individuals are exposed to a sequence of input/output pairs [@deloshExtrapolationSineQua1997; @mcdanielEffectsSpacedMassed2013], or situations where observers are presented with a an incomplete scatterplot or line graph and make predictions about regions of the plot that don't contain data [@ciccioneCanHumansPerform2021; @courrieuQuickApproximationBivariate2012; @saidExtrapolationAccuracyUnderestimates2021;@schulzCommunicatingCompositionalPatterns2020].

@@

: **Experiment 1 - End of training performance**. The Intercept represents the average of the baseline (constant condition), and the conditVaried coefficient reflects the difference between the constant and varied groups. A larger positive estimates indicates a greater deviation (lower accuracy) for the varied group.  {.sm}

@@

::: {.content-visible unless-format="pdf"}
# References

::: {#refs}
:::

@@

::: {.content-visible unless-format="markdown"}
### Appendix - Project 1

{{< include Appendix/IGAS-Supplemental.qmd >}}

### Appendix - Project 2 - Experiment 1

{{< include Appendix/E1_Appendix.qmd >}}
:::

::: {.content-visible when-format="pdf"}

>>>>

---
title: Full Dissertation
subtitle: "The Role of Variability in Learning Transfer: A Similarity-Based Computational Approach"
#date: 04/04/24
author:
- name: Thomas Gorman
  affiliation: Indiana University
page-layout: full
code-fold: true
code-tools: true
toc: true
toc-location: body
toc-title: Contents
cache: true
execute: 
  warning: false
  eval: true
format:
  html: default
  hugo-md:
    echo: false
    include: true
    html-math-method: mathjax
    output-file: diss_full_hugo.md
  gfm:
    echo: false
    output: false
    output-file: diss_full_gfm.md
  hikmah-manuscript-pdf:
    echo: false
    output-file: "tg_full_disser.pdf"
    mainfont: "Linux Libertine O"
    mainfontoptions:
      - "Numbers=Proportional"
      - "Numbers=OldStyle"
    mathfont: "Libertinus Math"
prefer-html: true
---

@@

Early models of learning often assumed that discrete experiences with some task or category were not stored individually in memory, but instead promoted the formation of a summary representation, often referred to as a prototype or schema, and that exposure to novel examples would then prompt the retrieval of whichever preexisting prototype was most similar. In addition to being a landmark study on the influence of training variability, @posnerGenesisAbstractIdeas1968 (described above) also put forward an influential argument concerning the nature of the mental representations acquired during learning - namely that learners tend to abstract a prototype, or aggregate representation of the dot pattern categories, rather than encoding each individual stimuli. Recall that participants are trained on only on distortions of the category prototypes (e.g.  low, medium or high distortions), never encountering the exact prototypes during the training stage. Then, in the testing phase, participants are tested with the prototype patterns, their old training items, and novel low, medium and high distortions. The authors found that participants had the highest testing accuracy for the previously unseen prototype patterns, followed by the old training items, and then the novel low, medium and high distortions. The authors interpreted this pattern as evidence that participants had acquired prototype representation of the category, as opposed to storing each individual training instance, and that generalization was based on the similarity of the testing items to the learned prototype representations. @posnerGenesisAbstractIdeas1968 has been extremely influential,  and continues to be cited in contemporary research as clear evidence that prototype abstraction underlies the benefits of varied training. It's also referenced as a key influence in the development of "Schema Theory of Motor Learning" @schmidtSchemaTheoryDiscrete1975, which in turn influenced decades of research on the potential benefits of varied training in motor skill learning. However a number of the core assumptions utilized by @posnerGenesisAbstractIdeas1968 were later called into question both empirically and with competing theoretical accounts [@hintzmanMINERVASimulationModel1984; @hintzmanSchemaAbstractionMultipletrace1986; @palmeriCentralTendenciesExtreme2001; @zakiHighdistortionEnhancementEffect2007; @mcclellandDistributedMemoryRepresentation1985; @nosofskyInvestigationsExemplarBasedConnectionist1992; @knappTheoryCategorizationBased1984]. @palmeriCentralTendenciesExtreme2001 demonstrated both the dangers of assuming that psychological representations mimic the metric stimulus space, as well the viability of models with simpler representational assumptions. These authors conducted a near replication of the @posnerGenesisAbstractIdeas1968 study, but also had participants provide similarity judgements of the dot pattern stimuli after completing the training phase. A multidimensional scaling analysis of the similarity judgements revelead that the psychological representations of the prototype stimuli were not located in the middle of the training stimuli, but were instead extreme points in the psychological space. The authors also demonstrated the generalization patterns of @posnerGenesisAbstractIdeas1968 could be accounted for by an exemplar-based model, without any need to assume the abstraction of a prototype.

@@

@@

The study of human function learning investigates how people learn relationships between continuous input and output values.  Function learning is studied both in tasks where individuals are exposed to a sequence of input/output pairs [@deloshExtrapolationSineQua1997; @mcdanielEffectsSpacedMassed2013], or situations where observers are presented with an incomplete scatterplot or line graph and make predictions about regions of the plot that don't contain data [@ciccioneCanHumansPerform2021; @courrieuQuickApproximationBivariate2012; @saidExtrapolationAccuracyUnderestimates2021;@schulzCommunicatingCompositionalPatterns2020].

@@

: **Experiment 1 - End of training performance**. The Intercept represents the average of the baseline (constant condition), and the conditVaried coefficient reflects the difference between the constant and varied groups. A larger positive estimate indicates a greater deviation (lower accuracy) for the varied group.  {.sm}

@@

@@

[Apppendix](Appendix/Full_Appendix.qmd) available at tegorman13.github.io/Dissertation/Sections/Appendix.html

### 2024-04-12 Sections/hold_temp.qmd
adding more citations to intro
update/re-render all versions. Took authors note out of hikmah manuscript

@@

>>>>

In North 2019 - biggest effect for varied-random. 
North, J. S., Bezodis, N. E., Murphy, C. P., Runswick, O. R., Pocock, C., & Roca, A. (2019). The effect of consistent and varied follow-through practice schedules on learning a table tennis backhand. Journal of Sports Sciences, 37(6), 613–620. https://doi.org/10.1080/02640414.2018.1522683

@@

```{md}
---
title: Full Dissertation
subtitle: "The Role of Variability in Learning Transfer: A Similarity-Based Computational Approach"
date: 04/04/24
author:
- name: Thomas Gorman
  affiliation: Indiana University
page-layout: full
code-fold: true
code-tools: true
toc: true
toc-location: body
toc-title: Contents
cache: true
execute: 
  warning: false
  eval: true
format:
  html: default
  hugo-md:
    echo: false
    #include: false
    html-math-method: mathjax
    output-file: diss_full_hugo.md
  gfm:
    echo: false
    output: false
    output-file: diss_full_gfm.md
  hikmah-manuscript-pdf:
    echo: false
    output-file: "tg_full_disser.pdf"
    mainfont: "Linux Libertine O"
    mainfontoptions:
      - "Numbers=Proportional"
      - "Numbers=OldStyle"
    mathfont: "Libertinus Math"
# editor: 
#   markdown: 
#     wrap: 72
prefer-html: true
---

::: {.content-visible when-format="html"}

{{< include pre_sections/pre_html.qmd >}}

:::

::: {.content-visible when-format="pdf"}

{{< include pre_sections/pre_pdf.qmd >}}

:::

::: {.content-visible when-format="docx"}
{{< include pre_sections/pre_docx.qmd >}}
:::

{{< pagebreak >}}

::: column-page-inset-right

# Abstract

The past century of research on human learning has produced ample evidence that although learners can improve at almost any task, such improvements are often specific to the trained task, with unreliable or even nonexistent transfer or generalization to novel tasks or conditions [@barnettWhenWhereWe2002; @dettermanCaseProsecutionTransfer1993]. Such generalization challenges are of noteworthy practical relevance, given that educators, trainers, and rehabilitators typically intend for their students to be able to apply what they have learned to new situations. It is therefore important to better understand the factors that influence generalization, and to develop cognitive models that can predict when generalization is likely to occur. Such characteristics have included training difficulty, spacing, temporal order, feedback schedules, and the primary focus of the current work - the variability of training examples.

This dissertation seeks to explore the cognitive underpinnings that govern the generalization of learning, focusing specifically on the role of variability during training in shaping subsequent transfer performance. A comprehensive review of the existing literature is presented, emphasizing the methodological complications associated with disentangling the confounding effects of similarity. Through a series of experiments involving several novel visuomotor tasks, this work investigates whether and how variability in training conditions affects performance in novel tasks. To theoretically account for the empirical outcomes, I employ both instance-based and connectionist computational models, both of which incorporate similarity-based mechanisms. These models serve to account for the extent to which variability influences the learners' generalization gradient, and also explain how training variation can produce both beneficial and deleterious outcomes.

{{< pagebreak >}}

::: {.content-visible when-format="html"}

{{< include pre_sections/toc_html.qmd >}}

:::

::: {.content-visible when-format="pdf"}

{{< include pre_sections/toc_pdf.qmd >}}

:::

{{< pagebreak >}}

\ \

::: {.content-visible unless-format="pdf"}
# References

::: {#refs}
:::

:::

# Appendix

[Apppendix](Appendix/Full_Appendix.qmd) available at tegorman13.github.io/Dissertation/Sections/Appendix.html

::: {.content-visible when-format="pdf"}

# References

:::

:::

<!-- ::: {.content-visible unless-format="markdown"}
### Appendix - Project 1

{{< include Appendix/IGAS-Supplemental.qmd >}}

{{< pagebreak >}}

### Appendix - Project 2 - Experiment 1

{{< include Appendix/E1_Appendix.qmd >}}
:::
 -->
```

### 2024-04-12 Sections/pre_sections/pre_docx.qmd
update/re-render all versions. Took authors note out of hikmah manuscript

Submitted to the faculty of the University Graduate School in partial fulfillment of the
requirements for the degree Doctor of Philosophy in the Department of Psychology and Brain
Sciences and the Cognitive Science Program, Indiana University.

@@

- Robert L. Goldstone, PhD
- Robert Nosofsky, PhD
- Peter Todd, PhD
- Mike Jones, PhD

>>>>

<!-- Submitted to the faculty of the University Graduate School in partial fulfillment of the
requirements for the degree Doctor of Philosophy in the Department of Psychology and Brain
Sciences and the Cognitive Science Program, Indiana University.

@@

- Robert L. Goldstone, PhD
- Robert Nosofsky, PhD
- Peter Todd, PhD
- Mike Jones, PhD -->

### 2024-04-12 Sections/pre_sections/pre_html.qmd
update/re-render all versions. Took authors note out of hikmah manuscript

<!-- Second Page and so on... -->
<div style="text-align: justify; margin-top: 700px;">
  Submitted to the faculty of the University Graduate School in partial fulfillment of the
  requirements for the degree Doctor of Philosophy in the Department of Psychology and Brain
  Sciences and the Cognitive Science Program, Indiana University
  Indiana University
</div>

@@

<!-- Committee Members Page -->
<div style="text-align: center; margin-top: 300px;">
  Accepted by the Graduate Faculty, Indiana University, in partial fulfillment of the
  requirements for the degree of Doctor of Philosophy.
</div>

@@

@@

<!-- Acknowledgements -->
<div style="text-align: center; font-weight: bold; font-size: 24px;">
  Acknowledgements
</div>
<div style="text-align: center; margin-top: 700px; font-size: 24px;">

</div>

>>>>

<!-- Second Page and so on... -->
<!-- <div style="text-align: justify; margin-top: 700px;">
  Submitted to the faculty of the University Graduate School in partial fulfillment of the
  requirements for the degree Doctor of Philosophy in the Department of Psychology and Brain
  Sciences and the Cognitive Science Program, Indiana University
  Indiana University
</div> -->

@@

<!-- Committee Members Page -->
<!-- <div style="text-align: center; margin-top: 300px;">
  Accepted by the Graduate Faculty, Indiana University, in partial fulfillment of the
  requirements for the degree of Doctor of Philosophy.
</div>

@@

<div style="page-break-before: always;"></div> -->

<!-- <div style="page-break-before: always;"></div> -->

<!-- Acknowledgements -->
<!-- <div style="text-align: center; font-weight: bold; font-size: 24px;">
  Acknowledgements
</div>
<div style="text-align: center; margin-top: 700px; font-size: 24px;">

</div> -->

@@

### 2024-04-12 Sections/pre_sections/pre_pdf.qmd
update/re-render all versions. Took authors note out of hikmah manuscript

Submitted to the faculty of the University Graduate School in partial fulfillment of the
requirements for the degree Doctor of Philosophy in the Department of Psychology and Brain
Sciences and the Cognitive Science Program, Indiana University
Indiana University

@@

\newpage

@@

\newpage

>>>>

<!-- Submitted to the faculty of the University Graduate School in partial fulfillment of the
requirements for the degree Doctor of Philosophy in the Department of Psychology and Brain
Sciences and the Cognitive Science Program, Indiana University
Indiana University -->

@@

<!-- \newpage

@@

\newpage -->
